---
title: "Quick reference: data.table"
author: "Taehoon Ha"
date: "7/15/2019"
output: 
    html_document:
        toc: TRUE
        toc_float:
            collapsed: TRUE
            smooth_scroll: FALSE
---

```{r setup, include=FALSE}
rm(list = ls())
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F, fig.align = "center")
library(tidyverse)
library(data.table)
library(lubridate)
```

### 1. Package loading

```{r}
library(data.table)
```

### 2. Example data: flights

```{r}
flights <- fread("Data/flights.csv")
head(flights) %>% knitr::kable()
```

### 3. What is `data.table`?

```{r}
class(flights)
```

> It is an extension of the `data.frame` class.

<br>

### 4. Subsetting a `data.table`

$$
DT[i, j, by] \\
\begin{aligned}
i&: \text{filters rows (analagous to a WHERE clause in SQL)} \\
j&: \text{is how you select/compute on columns} \\
by&: \text{is where you can group computations} \end{aligned}
$$

<br>

#### 1) $i$

+ first index is ALWAYS rows, regardless of commas

```{r}
identical(flights[2], flights[2,])
```

<br>

+ Compare to `data.frame`

```{r}
as.data.frame(flights)[2]  %>% dim
as.data.frame(flights)[2,] %>% dim
```

<br>

+ Use $i$ to select rows

```{r}
flights[1:2] # first and second row
flights[2:1] # second and first row
```

<br>

+ Logical expressions

```{r, eval = F}
flights[MONTH == 1]
flights[MONTH == 1 & DISTANCE > 1000]
```

<br>

+ `identical()` function

```{r}
identical(flights[-1:-100000], flights[!1:100000])
```

<br>

+ `order()`: optimized within a `data.table`

> + makes sorting a cinch
> + sort by one column and then another 
> + minus ('-') means descending order

```{r}
flights[order(TAIL_NUMBER,-ARRIVAL_DELAY)] %>% head() %>% knitr::kable()
```

<br>

> sort by a function on our data

```{r}
flights[order(substr(ORIGIN_AIRPORT,1,1))] %>% head %>% knitr::kable()
```

<br>

+ chain queries together

```{r}
flights[!is.na(ARRIVAL_DELAY)][order(TAIL_NUMBER,-ARRIVAL_DELAY)] %>% head() %>% knitr::kable()
```

<br>

```{r}
flights %>%
  .[!is.na(ARRIVAL_DELAY)] %>%
  .[order(TAIL_NUMBER,-ARRIVAL_DELAY)] %>% head() %>% knitr::kable()
```

<br>

#### 2) $j$

+ computations on columns

```{r}
# select columns by name in the second index parameter
flights[ , DESTINATION_AIRPORT] %>% head() %>% knitr::kable()           # as a vector
flights[ , list(DESTINATION_AIRPORT)] %>% head() %>% knitr::kable()     # as a data.table
```

<br>

+ use list to select multiple columns

```{r}
flights[ , list(DESTINATION_AIRPORT, ORIGIN_AIRPORT)] %>% head() %>% knitr::kable()
```

<br>

+ `.()` is the same as `list()` in `data.table`

```{r}
flights[ , .(DESTINATION_AIRPORT, ORIGIN_AIRPORT)] %>% head() %>% knitr::kable()
```

<br>

+ also computable on columns

```{r}
flights[ , unique(ORIGIN_AIRPORT)]  %>% head() %>% knitr::kable()
flights[ , sum(AIR_TIME, na.rm = TRUE) / (1e6 * 60)]  %>% head() %>% knitr::kable()
flights[ , summary(as.factor(CANCELLATION_REASON))]  %>% head() %>% knitr::kable()
flights[ , .(AIR_TIME, AIR_TIME/60)] %>% head() %>% knitr::kable()# return multiple columns
```

<br>

+ name the items

```{r}
flights[ , .(AIR_TIME, AIR_TIME_HOURS = AIR_TIME/60)] %>% head() %>% knitr::kable()
```

> Anytime the $j$ expression returns a list, the whole expression will return a `data.table`.

<br>

#### 3) $i$ and $j$

##### Q1. How many flights were delayed touching down?

```{r}
flights[ARRIVAL_DELAY>0, .N] 
```

> `.N` is a convenience symbol to get the number of rows in a `data.table`

<br>

##### Q2. What are some summary statistics of the delayed flights (in hours)?

```{r}
flights[ARRIVAL_DELAY>0, summary(ARRIVAL_DELAY/60)]
```

<br>

##### Q3. How many flights were on each day of the week?

```{r}
flights[ARRIVAL_DELAY>0, plot(table(DAY_OF_WEEK))] 
```

> + we don't have to return a `data.table`
> + an expression that returns a list in $j$ would, but here we want to plot, and `data.table` evaluates the expression correctly.

<br>

#### 4) `by`: grouping

##### Q4. How many outgoing flights did each aiport have?

```{r}
flights[ , .N, by = ORIGIN_AIRPORT] %>% head() %>% knitr::kable()
```

<br>

+ Order it to see the biggest airports

```{r}
flights[ , .N, by = ORIGIN_AIRPORT][order(-N)] %>% head() %>% knitr::kable()
```

<br>

##### Q5. Which airports had the most cancelled flights?

```{r}
flights[CANCELLED == 1, .N, by = ORIGIN_AIRPORT][order(-N)]
```

<br>

```{r eval = F}
flights %>%
  extract(CANCELLED == 1, .N, by = ORIGIN_AIRPORT) %>%
  extract(order(-N))
# Here we use extract, magrittr's alias for [ rather than the .[] notation above
```

<br>

##### Q6. What's the min and max departure delay for each airport, sorted by biggest difference between the two?

```{r}
flights[ , .(MinDelay = min(DEPARTURE_DELAY, na.rm = TRUE), 
             MaxDelay = max(DEPARTURE_DELAY, na.rm = TRUE)), 
         by = ORIGIN_AIRPORT][order(MinDelay-MaxDelay)] %>% head() %>% knitr::kable()
```

<br>

#### 5) `.SD`: a special name for the subset of the data.table for each group

```{r}
mtcarsDT <- as.data.table(mtcars, keep.rownames = 'Model')
mtcarsDT[ ,unique(cyl)] # we have three groups by cylinder count
mtcarsDT[ , print(.SD), by = cyl] # so we have three data.tables that get printed
```

> Ref: https://www.youtube.com/watch?v=L6aT_oEhIKo

<br>

##### Q7. First two rows of flights from each day

```{r}
flights[ , .SD[1:2], by = .(YEAR,MONTH,DAY)] %>% head() %>% knitr::kable()
```

<br>

Q8. 100 random obs from every day

```{r}
flights[ , .SD[sample(1:.N, 100)], by = .(YEAR, MONTH,DAY)] %>% head() %>% knitr::kable()
```

<br>

##### Q8. % of flights cancelled by airport?

```{r}
flights[ , .(PctCancelled = mean(CANCELLED) %>% round(., 3)), by = ORIGIN_AIRPORT][order(-PctCancelled)] %>% head() %>% knitr::kable()
```

<br>

##### Q9. Remove all airports from the dataset that don't have 10k flights

```{r}
flights[ , .SD[.N >= 10000], by = ORIGIN_AIRPORT] %>% head() %>% knitr::kable()
```

> + Here we're saying to group by the airport of origin
> + So `.SD` in each case is the complete set of flights for an airport
> + We then filter that subset. We put in a single `TRUE` or `FALSE`, is the number of rows greater than 10000? 
> + If it is, it returns TRUE and is recycled so that every row in the subset is returned. 
> + If not, then the FALSE is returned, recycled for all rows, and no rows are returned.
> + `data.table` will build up the final dataset based on these little data.tables
> + Appending zero rows to a data.table means nothing, and the airports that don't have enough rows are dropped from the final output.

<br>

##### Q10. Which airports don't have 10000 flights?

```{r}
flights[ , .SD[.N < 10000], by = ORIGIN_AIRPORT][ , unique(ORIGIN_AIRPORT)] %>% head()
```

<br>

#### 6) Assignment

> + there is an assignment operator that *modifies in place* (by reference)
> + jargon translation: it will alter the specific object you use it on

<br>

+ Create a date column that's in date format

```{r}
flights[ , Date := ymd(YEAR*10000+MONTH*100+DAY)] %>% head() %>% knitr::kable()
```

<br>

+ Reformat the dataset into boolean format

```{r}
flights[ , CANCELLED := ifelse(CANCELLED == 1, TRUE, FALSE)]
```

<br>

+ Create new columns

```{r}
flights[ , AvgSpeed := DISTANCE / (AIR_TIME/60)]
```

<br>

+ Update the `data.table` selectively

```{r eval = F}
flights[ CANCELLATION_REASON2 == 'A', CANCELLATION_REASON := 'Airline' ]
flights[ CANCELLATION_REASON2 == 'B', CANCELLATION_REASON := 'Weather'] 
flights[ CANCELLATION_REASON2 == 'C', CANCELLATION_REASON := 'National Air System'] 
flights[ CANCELLATION_REASON2 == 'D', CANCELLATION_REASON := 'Security'] 
flights[ CANCELLATION_REASON2 == '', CANCELLATION_REASON := NA] 
```

```{r eval = F}
flights[!is.na(CANCELLATION_REASON2), .N, by = CANCELLATION_REASON2] %>%
    ggplot(aes(x = CANCELLATION_REASON2, y = N)) +
    geom_histogram(stat = 'identity')
```

<br>

```
[As a function]

cancelReasonUpdate <- function(inStr){
  switch(inStr,
         "A" = "Airline",
         "B" = "Weather",
         "C" = "National Air System",
         "D" = "Security",
               NA)
}

flights[ , CANCELLATION_REASON2 := map_chr(CANCELLATION_REASON,cancelReasonUpdate)]
```

<br>

+ Delete columns

```{r eval = F}
flights[ , CANCELLATION_REASON2 := NULL]
```

<br>

+ Grouping

```{r}
flights[ , MaxAvgSpeed := max(AvgSpeed, na.rm = TRUE), by = .(ORIGIN_AIRPORT, DESTINATION_AIRPORT)]
```

> note that the group level statistic is now added to each observation

<br>

#### 6) `join`

+ Load the `airports` dataset

```{r}
airports <- fread("Data/airports.csv")
airports %>% head() %>% knitr::kable()
```

<br>

##### a. inner join

```{r}
merge(x = flights,
      y = airports,
      by.x = 'ORIGIN_AIRPORT',
      by.y = 'IATA_CODE') %>% head() %>% knitr::kable()
```

<br>

##### b. left outer join 

```{r}
flights[airports, on = c(ORIGIN_AIRPORT = 'IATA_CODE')] %>% head() %>% knitr::kable()
flights[airports, on = "ORIGIN_AIRPORT == IATA_CODE"] %>% head() %>% knitr::kable()
```

<br>

##### c. inner join

```{r}
flights[airports, on = c(ORIGIN_AIRPORT = 'IATA_CODE'), nomatch = 0] %>% head() %>% knitr::kable()
```

<br>

##### d. not join (a.k.a anti-join)

```{r}
airports[!flights, on = "IATA_CODE == ORIGIN_AIRPORT"] %>% head() %>% knitr::kable()
```

<br>

***

+ Ref: Cheat sheet of `data.table`: [Click]("https://github.com/rstudio/cheatsheets/raw/master/datatable.pdf")

<br>

***

<center><strong>End of Document</strong></center>


