---
title: "Monte Carlo Simulation for Causal Inference (G-computation formula)"
output:
  html_document:
    toc: true
    toc_depth: 5
    toc_float:
     collapsed: false
     smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

### A specific data generating process

In this simulation, we will generate the observed data O in the following way.

1. Draw $W1$ from a Bernoulli($p = 0.20$)

2. Given $W1$, draw $W2$ from a $Bernoulli(p = expit(0.5 * W1))$

3. Given $W1$, $W2$ draw $A$ from a $Bernoulli(p = expit[W1 * W2])$

4. Given $W1$, $W2$, $A$ draw $C$ from a $Bernoulli(p = expit[-A + 0.3 * W1 - W2])$

5. Given $W1$, $W2$, $A$, $C$, draw $Y$ from a $Normal(\mu = (4.4 * A + 0.7 * W1 - 2 *A * W2 - 2 * A * C), \sigma = 0.3^2)$ The random errors are independent.

<br>

#### 1. Use Monte Carlo simulation to evaluate the true value of the following parameter of the observed data distribution : 

$$
\phi = E_{0,W1,W2,C} [ \mathrm{E}(Y|A=1, W1, W2, C) - \mathrm{E}(Y|A=0, W1, W2, C)]
$$


```{r}
n <- 5000
R <- 10000 # 10K simulations
phi <- rep(NA, R)

for (r in 1:R) {
  W1 <- rbinom(n, 1, prob=0.20)
  W2 <- rbinom(n, 1, prob=plogis(0.5*W1))
  A <- rbinom(n, 1,  prob=plogis(W1*W2))
  C <- rbinom(n, 1, prob=plogis(-A+0.3*W1-W2))
  Y <- rnorm(n=n, mean = 4.4*A + 0.7*W1 - 2*A*W2 - 2*A*C, sd = 0.3)
  
  Y_1 <- 4.4 * 1 + 0.7 * W1 - 2 * 1 * W2 - 2 * 1 * C
  Y_0 <- 4.4 * 0 + 0.7 * W1 - 2 * 0 * W2 - 2 * 0 * C
  phi[r] <- mean(Y_1 - Y_0)
}

mean(phi)
```

<br>

> Based on 10000 Monte Carlo simulations, the true value of the following parameter of the observed data distribution, was `r mean(phi) %>% round(3)`. 
> $\phi = E_{0,W1,W2,C} [ \mathrm{E}(Y|A=1, W1, W2, C) - \mathrm{E}(Y|A=0, W1, W2, C)] =$ `r mean(phi) %>% round(3)`.

<br>

#### 2. Use Monte Carlo simulation to evaluate the true value of the average treatment effect: 
$$
\theta^{*} = \mathrm{E}^*(Y_1)- \mathrm{E}^*(Y_0)
$$

```{r}
estimates <- rep(NA, R)

for (r in 1:R) {
  W1 <- rbinom(n, 1, prob=0.20)
  W2 <- rbinom(n, 1, prob=plogis(0.5*W1))
  A <- rbinom(n, 1,  prob=plogis(W1*W2))
  C <- rbinom(n, 1, prob=plogis(-A+0.3*W1-W2))
  Y <- rnorm(n=n, mean = 4.4*A + 0.7*W1 - 2*A*W2 - 2*A*C, sd = 0.3)
  
    
  Obs <- data.frame(W1, W2, A, C, Y)
    
  txt <- control <- Obs
  
  txt$A <- 1
  control$A <- 0
  
  reg1 <- glm(Y ~ A + W1 + W2 + C, data=Obs)
  
  # reg1
  pred1.txt <- predict(reg1, newdata = txt)
  pred1.ctrl <- predict(reg1, newdata = control)
    
  theta_star <- mean(pred1.txt-pred1.ctrl)
    
  estimates[r] <- theta_star
}

mean(estimates)
```

<br>

> Based on 10000 Monte Carlo simulations, the true value of the average treatment effect, was `r mean(estimates) %>% round(3)`. $\theta^{*}(P^{*}) = `r mean(estimates) %>% round(3)`$ is the difference in the counterfactual expected weight gain if all children were given RUTF and the counterfactual expected weight gain if all children were given the standard supplement.

<br>

***

<center><strong>End of Document</strong></center>